---
title: "Final"
output:
  pdf_document: default
  html_document: default
date: "2024-04-24"
---

```{r}


#install.packages("readxl")
library(readxl)
library(ggplot2)
setwd("~/Desktop")

price_data <- read_excel("Tomato Prices.xls")
price_data

price_data_df <- as.data.frame(price_data)


colnames(price_data_df) <- c("date", "price")

orig_data_df <- price_data_df




price_data <- ts(price_data_df$price)  
plot(price_data, xlab = "Months from January 1st, 1980", ylab = "price", main = "Average Price of Tomatoes per Pound Across US Cities")




```

# Examine Data for Outliers and Clean with tsclean():

```{r}


library(forecast)

adjusted_series <- tsclean(price_data)

plot(adjusted_series, type = "l", xlab = "Time", ylab = "Value", main = "Adjusted Time Series Data")

adjusted_df <- data.frame(date = seq_along(adjusted_series), price = adjusted_series)

print(adjusted_df)

price_data_df <- adjusted_df
orig_data_df <- adjusted_df

plot(price_data_df$date, price_data_df$price, type = "l", xlab = "Date", ylab = "Price", main = "Cleaned Time Series Data")

```

```{r}


#price_data_df$date <- as.numeric(price_data_df$date)
#price_data_df$price <- as.numeric(price_data_df$price)



#install.packages("VIM")
#library(VIM)

#df_imputed <- kNN(price_data_df, k = 5)

#price_data_df$price <- df_imputed$price



#price_data_df$date <- orig_data_df$date


#price_data <- ts(price_data_df$price)  


#price_data_df

#price_data <- ts(price_data_df$price)  
#plot(price_data, xlab = "Months from January 1st, 1980", ylab = "price", main = "Average US City Tomato Price")
```

# Classical Decomposition

```{r}



ts_data <- ts(price_data_df$price, frequency = 12)  

ts_data_original <- ts_data


origin_data <- ts(orig_data_df$price, frequency = 12)


decomposed_ts <- decompose(ts_data, type = "additive")  




plot(decomposed_ts$trend, main = "Trend", xlab="Date in Months from January 1st 1980", ylab = "Price in Dollars")

plot(decomposed_ts$seasonal, main = "Season", xlab = "Date in Months from January 1st 1980", ylab = "Price in Dollars")

plot(decomposed_ts$random, main = "Random", xlab="Date in Months from January 1st 1980", ylab = "Price in Dollars")



par(mfrow = c(3, 1)) 

plot(decomposed_ts$trend, main = "Trend", xlab="Date", ylab = "price")

plot(decomposed_ts$seasonal, main = "Season", xlab = "date", ylab = "price")

plot(decomposed_ts$random, main = "Random", xlab="date", ylab = "price")


```

# Dickey-Fuller Test to Check for Unit Root:

```{r}


#install.packages("tseries")
library(tseries)

ts_values<-as.numeric(origin_data)

adf_test_result<-adf.test(ts_values)

print(adf_test_result)

```

Since the p-value is greater than .05, we accept the Null and conclude that the data is non-stationary

# Conduct Differencing and Seasonal Differencing:

```{r}

trend_removed_ts<-diff(ts_data, differences = 1)

seasonal_diff_ts <- diff(trend_removed_ts, lag = 365)

plot(seasonal_diff_ts, main = "Price Data without trend and seasonal components", xlab = "date", ylab = "diff(1), diff(12)")

seasonal_diff_ts_numeric <- as.numeric(seasonal_diff_ts)


```

# Dickey-Fuller Test to Check for Unit Root:

```{r}

#install.packages("tseries")
library(tseries)

seasonal_diff_ts_numeric <-as.numeric(seasonal_diff_ts_numeric)

adf_test_result <-adf.test(seasonal_diff_ts_numeric)

print(adf_test_result)




```

Since the p-value is less than .05, we conclude that the data is stationary

# ACF and PACF of stationary data - (after 1 differencing and 1 seasonal differencing):

```{r}

acf(seasonal_diff_ts_numeric, main="ACF")
pacf(seasonal_diff_ts_numeric, main="PACF")



```

# Possible ARIMA and SARIMA Models from ACF and PACF plots - (fit to original data):

```{r}

library(forecast)



arima_models <- list()

arima_model_001 <- arima(ts_data_original, order = c(0, 1, 0), method = "ML")
arima_models[["ARIMA(0,1,0)"]] <- arima_model_001

arima_model_011 <- arima(ts_data_original, order = c(0, 1, 1), method = "ML")
arima_models[["ARIMA(0,1,1)"]] <- arima_model_011

arima_model_012 <- arima(ts_data_original, order = c(0, 1, 2), method = "ML")
arima_models[["ARIMA(0,1,2)"]] <- arima_model_012

arima_model_101 <- arima(ts_data_original, order = c(1, 1, 0), method = "ML")
arima_models[["ARIMA(1,1,0)"]] <- arima_model_101

arima_model_111 <- arima(ts_data_original, order = c(1, 1, 1), method = "ML")
arima_models[["ARIMA(1,1,1)"]] <- arima_model_111

arima_model_112 <- arima(ts_data_original, order = c(1, 1, 2), method = "ML")
arima_models[["ARIMA(1,1,2)"]] <- arima_model_112

arima_model_201 <- arima(ts_data_original, order = c(2, 1, 0), method = "ML")
arima_models[["ARIMA(2,1,0)"]] <- arima_model_201

arima_model_211 <- arima(ts_data_original, order = c(2, 1, 1), method = "ML")
arima_models[["ARIMA(2,1,1)"]] <- arima_model_211

arima_model_212 <- arima(ts_data_original, order = c(2, 1, 2), method = "ML")
arima_models[["ARIMA(2,1,2)"]] <- arima_model_212





sarima_models <- list()

sarima_model_0010010 <- arima(ts_data_original, order = c(0, 1, 0), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(0,1,0)(0,1,0)10"]] <- sarima_model_0010010

sarima_model_0110010 <- arima(ts_data_original, order = c(0, 1, 1), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(0,1,1)(0,1,0)10"]] <- sarima_model_0110010

sarima_model_0120010 <- arima(ts_data_original, order = c(0, 1, 2), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(0,1,2)(0,1,0)10"]] <- sarima_model_0120010

sarima_model_1010010 <- arima(ts_data_original, order = c(1, 1, 0), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(1,1,0)(0,1,0)10"]] <- sarima_model_1010010

sarima_model_1110010 <- arima(ts_data_original, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(1,1,1)(0,1,0)10"]] <- sarima_model_1110010

sarima_model_1120010 <- arima(ts_data_original, order = c(1, 1, 2), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(1,1,2)(0,1,0)10"]] <- sarima_model_1120010

sarima_model_2010010 <- arima(ts_data_original, order = c(2, 1, 0), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(2,1,0)(0,1,0)10"]] <- sarima_model_2010010

sarima_model_2110010 <- arima(ts_data_original, order = c(2, 1, 1), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(2,1,1)(0,1,0)10"]] <- sarima_model_2110010

sarima_model_2120010 <- arima(ts_data_original, order = c(2, 1, 2), seasonal = list(order = c(0, 1, 0), period = 10), method = "ML")
sarima_models[["SARIMA(2,1,2)(0,1,0)10"]] <- sarima_model_2120010



print("ARIMA Models:")
print(arima_models)

print("SARIMA Models:")
print(sarima_models)


```

# ARIMA Forecast with Best Model:

```{r}

library(forecast)

forecast_values <- forecast(arima_model_212, h = 20)

autoplot(forecast_values)


```

# Model Diagnostics for Best ARIMA Model:

```{r}

plot(arima_model_212$residuals / sqrt(arima_model_212$sigma2))


qqnorm(residuals(arima_model_212))
qqline(residuals(arima_model_212))


acf(residuals(arima_model_212))

ljung_box_results <- Box.test(residuals(arima_model_212), lag = 20, type = "Ljung-Box")
print("Ljung-Box Test Results:")
print(ljung_box_results)



```

# Model the Data with Auto.ARIMA function to compare:

```{r}


library(forecast)

model <- auto.arima(ts_data_original, seasonal = TRUE)

summary(model)


forecast_result <- forecast(model, h = 50)
print(forecast_result)
plot(forecast_result)
```

# Model Diagnostics for Best auto Arima Model: 

```{r}

plot(model$residuals / sqrt(model$sigma2))


qqnorm(residuals(model))
qqline(residuals(model))


acf(residuals(model))

ljung_box_results <- Box.test(residuals(model), lag = 20, type = "Ljung-Box")
print("Ljung-Box Test Results:")
print(ljung_box_results)



```

# Smoothing-based Forecast: Holt-Winters:

```{r}


library(forecast)


holtwintersmodel <- hw(ts_data_original, seasonal = "additive")

forecast_values <- forecast(holtwintersmodel, h = 20)

print(forecast_values)
plot(forecast_values)



```

# 

# Evaluation of Forecast Accuracy - SARIMA vs. Holt-Winters - (MAE, RMSE, MAPE) 

Divide dataset into test and train

```{r}

testlen <- round(length(ts_data_original) * 0.1)

train_data <- head(ts_data_original, -testlen)
test_data <- tail(ts_data_original, testlen)

```

Refit the ARIMA models to the new training data;

```{r}

arima_model <- auto.arima(train_data, seasonal = TRUE)
arima_forecast <- forecast(arima_model, h = 20)


```

Refit Holt-Winters to new data:

```{r}

holtwinters_model <- hw(train_data, seasonal = "additive")

holtwinters_forecast <- forecast(holtwinters_model, h = 20 )


```

Create Function to get MAE, RMSE, MAP

```{r}

forecast_accuracy <- function(forc, act){
  
  

    errors <- forc-act
    
    
    mae <- mean(abs(errors))
    
    
    rmse <- sqrt(mean(errors^2))
    mape <- mean(abs(errors/  act))*100
    
    
    return(c(MAE=mae, RMSE = rmse, MAPE = mape))
    
    
    
    
    
}

```

```{r}

arima_acc <- forecast_accuracy(arima_forecast$mean, test_data)


print("ARIMA Acc:")
print(arima_acc)

```

```{r}


HW_acc <- forecast_accuracy(holtwinters_forecast$mean, test_data)
print("Holt-Winter Acc:")
print(HW_acc)


```

The ARMIA forecast has lower MAE, RMSE and MAPE and is the better forecast.

# 
